\documentclass[11pt,letterpaper]{article}

% XeLaTeX for Turkish character support
\usepackage{fontspec}
\setmainfont{DejaVu Serif}
\setsansfont{DejaVu Sans}

% Page layout
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Colors and styling
\usepackage[dvipsnames]{xcolor}
\definecolor{bentleyblue}{RGB}{0, 51, 102}
\definecolor{linkblue}{RGB}{0, 102, 204}

% Headers and sections
\usepackage{titlesec}
\titleformat{\section}{\Large\bfseries\color{bentleyblue}}{\thesection}{1em}{}[\titlerule]
\titleformat{\subsection}{\large\bfseries\color{bentleyblue!80}}{\thesubsection}{1em}{}
\titleformat{\subsubsection}{\normalsize\bfseries\color{bentleyblue!60}}{\thesubsubsection}{1em}{}

% Lists
\usepackage{enumitem}
\setlist[itemize]{leftmargin=*, itemsep=2pt, parsep=0pt}
\setlist[enumerate]{leftmargin=*, itemsep=2pt, parsep=0pt}

% Tables
\usepackage{booktabs}
\usepackage{tabularx}

% Links
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=linkblue,
    urlcolor=linkblue,
    pdfauthor={Onur Altındağ},
    pdftitle={EC224 Intermediate Microeconomics - Spring 2026 Syllabus}
}

% Header/Footer
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhf{}
\fancyhead[L]{\small\color{gray}EC 224 Intermediate Microeconomics}
\fancyhead[R]{\small\color{gray}Spring 2026}
\fancyfoot[C]{\thepage}
\renewcommand{\headrulewidth}{0.4pt}

% Remove paragraph indent, add space between paragraphs
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt}

\begin{document}

% Title
\begin{center}
{\LARGE\bfseries\color{bentleyblue} EC 224: Intermediate Microeconomics}\\[0.3cm]
{\Large Spring 2026 Syllabus}\\[0.5cm]
{\large Bentley University}
\end{center}

\vspace{0.5cm}

%-----------------------------------------------------------
\section*{Course Description}
%-----------------------------------------------------------

This is an intermediate microeconomics course that provides an analytical approach to economic interactions between main economic actors (individuals, firms, and governments). The course builds on the concepts you learned in introduction to microeconomics and more formally studies the price mechanism, production, and economic behavior in different market settings and levels of information that actors have access to.

\subsection*{Knowledge}
\begin{itemize}
    \item Consumer maximization problem using utility functions, optimization of consumption subject to a budget constraint. Consumer responses to changes in market prices.
    \item Firm profit maximization using isoquant and isocost analysis subject to cost constraints. Firm responses to changes in market prices.
    \item Industries with different cost structures and returns to scale associated with them.
    \item Welfare, market failures, negative and positive externalities.
    \item Different market structures: perfect competition, monopolistic competition, oligopoly, monopoly.
\end{itemize}

\subsection*{Skills}
\begin{itemize}
    \item Advance analytical analysis and graphical reasoning using microeconomic theory.
    \item Use mathematical tools creatively to analyze economic problems.
    \item Develop a critical approach to economic theory and problem solving.
\end{itemize}

\subsection*{Perspectives}
\begin{itemize}
    \item Have a better understanding of market structure and the interaction of economic agents within the market.
    \item Have a better understanding of marginal decision-making approach to economic reasoning.
\end{itemize}

%-----------------------------------------------------------
\section*{Class Information}
%-----------------------------------------------------------

\begin{tabularx}{\textwidth}{@{}lX@{}}
\textbf{Instructor:} & Onur Altındağ \\
\textbf{Office:} & RAU 019F \\
\textbf{Email:} & \href{mailto:oaltindag@bentley.edu}{oaltindag@bentley.edu} \\
\textbf{Web:} & \url{https://www.onuraltindag.info/} \\[0.3cm]
\textbf{Course:} & EC 224-2 Intermediate Microeconomics \\
\textbf{Semester:} & Spring 2026 (January 20, 2026 – May 7, 2026) \\
\textbf{Meeting Time:} & Monday/Wednesday, 5:00 PM – 6:20 PM \\
\textbf{Location:} & Smith Technology Center 100 \\
\textbf{Delivery Mode:} & In-Person Lecture \\
\end{tabularx}

\vspace{0.3cm}

Office hours are held on Mondays and Wednesdays by appointment. Please \href{https://outlook.office.com/bookwithme/user/1c4e4f54e0f0493397e615b2c80a59c7\%40bentley.edu/meetingtype/PxtkOiMCu0m4QV32Y3tEaw2?bookingcode=097c05f0-ef41-47c9-9d76-fcf8bab2bc80&anonymous&ismsaljsauthenabled}{book a time slot} to schedule a meeting. Each appointment is 20 minutes maximum. If you need to meet urgently or no slots are available, email me directly.

%-----------------------------------------------------------
\section*{Important Dates and Evaluation}
%-----------------------------------------------------------

\begin{tabularx}{\textwidth}{@{}Xr@{}}
\toprule
\textbf{Component} & \textbf{Weight} \\
\midrule
Weekly assignments + participation & 25\% \\
\textbf{Monday, February 23, 2026} – First Midterm & 20\% \\
\textbf{Monday, April 6, 2026} – Second Midterm & 25\% \\
\textbf{Friday, May 1, 2026, 5:00 PM – 7:00 PM} – Final Exam & 30\% \\
\bottomrule
\end{tabularx}

%-----------------------------------------------------------
\section*{Grading}
%-----------------------------------------------------------

You \textbf{MUST} attend all the midterms and the final as there will be no make-up exams in this course. The midterms are \textbf{NOT} cumulative. If you miss or are likely to miss a midterm due to an emergency, please contact me as soon as possible. You will need to provide supporting documentation/verification of your absence. I will re-weight your final exam accordingly if you have a valid excuse. Please note that family vacations or missing the school shuttle are not valid excuses.

The final exam \textbf{is} cumulative. If you miss the final exam due to an emergency, you will receive an incomplete for this course. Do not take this class if you know that you will not be able to attend the final exam.

You have \textbf{10 homework assignments}. The submission system is automated and will not accept any submissions after the deadline (even 5 minutes). You will miss the assignment and there will be no exceptions. For each question, you have three attempts. If you receive at least \textbf{50\%} of the overall score indicated in the homework assignment, you will receive full credit. Otherwise, you will receive zero. You can miss up to 3 homework assignments; I will take the average of your best 7 homework scores as your homework grade. The questions are algorithmic and will be slightly different each time, so you are encouraged to work together on homework assignments.

%-----------------------------------------------------------
\section*{Registration on MindTap}
%-----------------------------------------------------------

\textbf{Course:} Section 1 of EC 224 Intermediate Microeconomics\\
\textbf{Instructor:} Onur Altındağ

This course requires an online learning platform called MindTap. Follow the instructions below to get started.

\subsection*{Register for your MindTap Course}
\begin{enumerate}
    \item Use the course registration link: \url{https://student.cengage.com/course-link/MTPNTBQGPQ8P}
    \item Follow the instructions on screen to create your Cengage account and register for this MindTap course.
    \item Begin your temporary access period.
\end{enumerate}

Need help? Visit the \href{https://startstrong.cengage.com}{Cengage Start Strong Website} for step-by-step instructions.

\textbf{Temporary Access:} You can access your MindTap course until 5:00 AM (UTC) on 2/3/2026 for free. At the end of the temporary access period, you will be prompted to purchase access. Your work will be saved and will be available to you again once you've completed your purchase.

\textbf{NOTE:} If the cost of your course materials is included in your tuition, you will not need to purchase access.

\subsection*{MindTap Tips \& Training Tools}
Learn more about navigating your MindTap course: \url{https://help.cengage.com/mindtap/mt-student/introduction.html}

\subsection*{Technical Support \& Troubleshooting}
Our US-based support team delivers answers and advice via 24/7 online chat, Twitter, live phone support (1-800-354-9706) and through \href{https://support.cengage.com}{support.cengage.com}, which includes helpful articles and tutorials.

If you are having trouble loading MindTap, run the \href{https://ng.cengage.com/static/browsercheck/index.html}{MindTap browser check} to make sure your browser is compatible or refer to the \href{https://help.cengage.com/mindtap/MindTap-System-Requirements.pdf}{MindTap System Requirements}. If MindTap isn't loading, be sure to visit \href{https://techcheck.cengage.com}{Techcheck} to see if there is an outage.

%-----------------------------------------------------------
\section*{Academic Integrity}
%-----------------------------------------------------------

Learning is a privilege that demands responsibility. At Bentley, students and faculty are members of an academic community that supports integrity both inside and outside the classroom. The expectation at Bentley is that students will take advantage of the opportunity for intellectual development and, in doing so, will conduct themselves in a manner consistent with the standards of academic integrity. When these standards are violated or compromised, individuals and the entire Bentley community suffer. Students who engage in acts of academic dishonesty not only face university censure but also may harm their future educational and employment opportunities. In other words, don't bring unauthorized materials into exams, don't plagiarize someone else's work, and make sure that your collaborations are conducted in accordance with university and course policy.

All students have access to Bentley's academic integrity policy in the \href{https://bentleydownloads.s3.amazonaws.com/registrar/Bentley+University+Student+Handbook+2025-2026.pdf}{Bentley University Student Handbook}. The best way to avoid a problem is to consult with your instructor before taking any action that might constitute a violation.

%-----------------------------------------------------------
\section*{Support Services}
%-----------------------------------------------------------

\subsection*{Bias Incident Response}
The Bias Incident Response Team (BIRT) provides students affected by bias or bias-related incidents with access to appropriate resources. Where appropriate, BIRT assists the University in its response to situations that may impact the overall campus climate related to diversity and inclusion. More information about BIRT and how to file a bias incident report can be found at: \url{https://www.bentley.edu/offices/student-affairs/birt}

\subsection*{Student Accessibility Services}
Bentley University abides by Section 504 of the Rehabilitation Act of 1973 and the Americans with Disabilities Act of 1990 which stipulate no student shall be denied the benefits of an education solely by reason of a disability. If you have a hidden or visible disability which may require classroom accommodations, please call Student Accessibility Services within the first 4 weeks of the semester to schedule an appointment. Student Accessibility Services is located in the Office of Student Success (JEN 336, 781.891.2004). Student Accessibility Services is responsible for managing academic accommodations and services for all students with disabilities.

For more information, visit: \url{https://www.bentley.edu/offices/student-accessibility-services}

The Undergraduate Academic Services (UAS) Peer Tutoring program offers online one-on-one and small group tutoring services for students who have worked with their instructors and made use of the Learning Centers, but still require additional academic support. Please reach out to me if you need more information.

%-----------------------------------------------------------
\section*{Attendance Policy}
%-----------------------------------------------------------

All students must attend the in-person classes. In-person attendance is essential because it allows for real-time interaction, immediate feedback, and the kind of dialogue that makes learning stick. I do not record lectures, so if you miss a class, you miss the material.

If you have an exceptional circumstance that prevents you from attending in person, you must notify me \textbf{before} class. If I approve, I will provide you with a Zoom link to join remotely. Remote attendance is a privilege, not a default option.

There may be one or two sessions during the semester when I need to hold class online due to my schedule. I will notify you in advance when this happens.

%-----------------------------------------------------------
\section*{Tentative Schedule}
%-----------------------------------------------------------

\begin{itemize}
    \item \textbf{Week 1:} Utility and Choice – Ch. 2
    \item \textbf{Week 2:} Utility and Choice – Ch. 2 \& Demand Curves – Ch. 3
    \item \textbf{Week 3:} Demand Curves – Ch. 3 \& Uncertainty – Ch. 4
    \item \textbf{Week 4:} Uncertainty – Ch. 4
    \item \textbf{Week 5:} Game Theory – Ch. 5
    \item \textbf{Week 6:} Game Theory – Ch. 5 / \textbf{First Midterm (Feb 23)}
    \item \textbf{Week 7:} Production – Ch. 6
    \item \textbf{Week 8:} Production – Ch. 6
    \item \textbf{Week 9:} Costs – Ch. 7
    \item \textbf{Week 10:} Profit Maximization and Supply – Ch. 8 / \textbf{Second Midterm (Apr 6)}
    \item \textbf{Week 11:} Perfect Competition in a Single Market – Ch. 9
    \item \textbf{Week 12:} Monopoly – Ch. 11
    \item \textbf{Week 13:} Imperfect Competition – Ch. 12
    \item \textbf{Week 14:} Imperfect Competition – Ch. 12
    \item \textbf{Week 15:} Review and Final Exam
\end{itemize}

\textit{Note: Schedule is subject to change. Chapter references are from Nicholson and Snyder's Microeconomic Theory.}

\end{document}
